Regional Effects of Marine Protected Area Networks Are Unclear

### 2020-04-19 documents/ovando-regional-effects-of-mpas-v2.Rmd
confirmed knit with updated edits, turning to cover letter

---
title: Regional Effects of Marine Protected Area Networks Are Unclear
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity called for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature is now calling for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they may benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of numerical or biomass densities inside their borders, relative to these metrics in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of this challenge. A network of MPAs covering 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. We are unable to find a clear effect of these same MPAs on biomass densities throughout the Channel Islands region. We use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::word_document2: default
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

#   sim_plot <- outcomes %>%
#     mutate(rsize = plyr::round_any(mpa_size, 0.25),
#            reff = plyr::round_any(mpa_effect,0.2)) %>%
#     filter(rsize == 0.25, depletion < 0.8, depletion > 0.2,
#            scientific_name %in% sim_life_history$scientific_name
#            ) %>%
#     mutate(year = years_protected + 2001) %>%
#     filter(year %in% did_betas$year, years_protected >1) %>% 
#     group_by(year, reff) %>% 
#     count() %>% 
#     group_by(year) %>% 
#     mutate(sc = n / sum(n))
# 
#   did_plot <- did_betas %>%
#     ggplot() +
#     geom_tile(data = sim_plot, aes(year, reff, fill = sc), alpha = 0.75) +
#     geom_vline(aes(xintercept = 2003), color = 'red', linetype = 2, size = 2) +
#     geom_hline(aes(yintercept = 0)) +
#     geom_pointrange(aes(
#       year,
#       y = exp(estimate) - 1,
#       ymin = exp(lower) - 1,
#       ymax = exp(upper) - 1
#     ),
#     size = 1.5
#     ) +
#     labs(x = "Year", y = "Targeted Trend Divergence") +
#     scale_fill_viridis(guide = wide_hgc,
#                        name = "% of Annual Sims",
#                        labels = scales::percent_format(accuracy = 1),
#                        option = "magma") + 
#     theme(legend.position = "top",
#         panel.grid.major.x = element_blank(),
#         panel.grid.minor.x = element_blank()
#         )
#   
# 
# 
# 
# cip_data <- pisco_data %>%
#   left_join(site_data, by = c("site","side")) %>%
#   filter(is.na(eventual_mpa) == F) %>%
#   filter(region %in% c("ANA", "SCI","SRI",'SMI'),
#          classcode %in% top_species)

# process outcomes

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas, parts of the oceans protected from human activities such as fishing, are increasingly being used to conserve and manage these ecosystems. Our study pairs theory with empirical methods to examine the regional effects of MPAs on conservation outcomes. We show that the regional effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a network of MPAs after 14 years of protection, which our simulation analysis suggests is actually to be expected. Our results lay a foundation for future research on the design and monitoring of marine protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years. DO thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments and technical support and special thanks to Kathyrn Davis Koehn and Avrey Parsons-Field.

@@

Many recent agreements to expand MPAs (most proximately, the Convention on Biological Diversity’s Strategic Plan for Biodiversity, which calls for 10% of coastal waters to be protected inside MPAs by 2020 and the International Union for Conservation of Nature call for 30% by 2030) are based on the expectation that well-designed MPAs will achieve benefits both within, and outside, their borders. Despite these assumptions, our collective scientific understanding of the regional-scale conservation and fishery impacts of current and future MPAs is surprisingly limited.

@@

As stakeholders around the world increasingly seek to use MPAs in marine resource management portfolios, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. This is particularly true in the context of a changing climate, in which "set it and forget it" management strategies are increasingly untenable; we need to be able to monitor the performance of MPAs in order to adapt them as dictated by a shifting environment. The MPAs within the Channel Islands National Marine Sanctuary, California, USA (which we will refer to as the Channel Islands from now on) provides an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 [see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs]. This MPA network has subsequently been used as a model case study in protected area design around the world [@botsford2014 and references therein].

@@

We define the regional conservation effects of MPAs as the change in mean total biomass densities of targeted finfish both inside and outside of MPAs, relative to the mean total biomass densities of targeted finfish inside and outside of MPAs that would have occurred without the MPAs (acknowledging that other outcomes such as increased biodiversity or resiliency are also important to conservation but are beyond the scope of this analysis). Consider a evenly distributed population that has 50% of its range protected by an MPA. Suppose that the MPA increase biomass densities inside the reserves by 20%, and by 0% outside the reserve. The regional effect of the MPA would be 10%.

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through habitat and network effects [@gaines2010; @costello2010a].

@@

The MPAs in this study cover 21% of the the surface waters of the Northern Channel Islands, a region spanning roughly 90km east to west. While detailed dispersal studies are not available for all of the species covered by our study, what information we have suggest that while adults of some of these species are likely to exhibit site fidelity [@love2011], larvae are likely distributed beyond the Channel Islands [@rassweiler2012]. Therefore, if we assume that the population of the targeted species is at least equal to the extent of the Channel Islands, we can assume that at most the MPAs cover roughly 20% of the targeted populations' ranges.

@@

We integrated these network design features, life history traits, and exploitation histories into a spatially explicit bio-economic simulation model, and used this model to generate expected outcomes for the Channel Islands MPAs. Our results suggest that while a wide range of outcomes are plausible, from 0% to upwards of 200%, across the majority of our simulations the effect size was 25\% or less, with a median simulated effect size of roughly 20\%; a potentially small value given the natural variability and sampling challenges of marine environments.

@@

In the case of the Channel Islands MPAs, control sites are often located within a few kilometers of an MPA, suggesting that they are susceptible to biological spillover and concentration of fishing effort resulting from the MPAs. Given these complications, we can certainly interpret the response ratios reported in @caselle2015, and updated here, as evidence that the MPAs are providing effective protection of targeted fish biomass inside their borders, particularly in light of the general increase in response ratios over time. However, it is unclear how well these response ratio results serve as in indicator of overall regional effects of the MPAs.

To illustrate this problem, we used our bio-economic model to simulate response ratio trajectories for species and MPA coverage representative of the Channel Islands (see methods and SI for details of model structure). These simulations cover the life-histories and MPA sizes seen in the Channel Islands, but vary in key unknowns such as the actual degree of fishing pressure, the timing of density dependence, and the fishing fleets response to the MPAs. For each year of protection, we paired our simulated response ratios to the estimated posterior probability distributions of the Channel Island response ratios. For example, if the mean estimated response ratio in the year 2006 is one, we found all simulations with three years of protection that had simulated response ratios near one, and then pulled out the "true" simulated regional MPA effect from each of those simulations. This provides us with a distribution of simulated regional MPA effects that could plausibly generate the types of response ratios actually observed in the Channel Islands.

As reported in @caselle2015, we see an increasing trend in response ratios of targeted species over time (Fig.\@ref(fig:rr-plot)). This provides strong evidence that the MPAs are providing protection for targeted species within their borders. However, our simulation results show that the response ratio trends we observe in the data could plausibly be produced by a wide range of regional MPA effects (Fig.\@ref(fig:rr-plot)). Response ratios well over one were associated with regional MPA effects generally less than 25%, and many simulations produced large response ratios but regional MPA effects close to 0% This can occur if for example fishing pressure is only moderate, adult movement is low, larval dispersal is high, and displaced fishing effort concentrates around the border of the MPAs.

@@

Building off of the concepts explored in @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group. Targeted species in the Channel Islands, include commercially important fin-fish such as California sheephead (*Semicossyphus pulcher*), and copper (*Sebastes caurinus*) and blue (*Sebastes mystinus*) rockfish. Each of these targeted species was the subject of prior bio-economic modeling related to the effects of MPAs in southern California [@costello2010a;@rassweiler2014]. It should be noted though that our analysis omits species from important invertebrate fisheries including red urchin (*Mesocentrotus franciscanus*) and spiny lobster (*Panulirus interruptus*). Included non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*) (See Table.S2 for a complete list of species). We used a Bayesian difference-in-difference regression to estimate any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation [see @larsen2019 for a succinct explanation of difference-in-difference estimators in ecological settings].  The result of this regression is, conditional on the assumptions of the model, an estimate of the effect of the MPAs on the mean total biomass densities of targeted species throughout the Channel Islands.

@@

As conventional wisdom and theory would suggest, over the first three years of implementation (2003-2006), the effects of the MPAs are unclear, with support for a small negative effect to a substantially positive effect, with much higher probability of a small positive effect (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`). Over the next six years the model estimates greater probabilities of an increasingly positive MPA effect, peaking in 2009-2011 with a median estimate of MPA effect of a `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)` increase in mean total biomass density of targeted species (90% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`. These empirical estimates are in line with (though in the upper range of) the outcomes that our simulation model suggests are plausible given the rough characteristics of the northern Channel Island MPAs. While this concordance between theory and empirics is by no means proof of the robustness of our results, it is an important line of evidence. However, in the subsequent years the trend reverses itself, and for the years 2015-2017 we once again see no clear effect of the MPAs (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`.) Fig.\@ref(fig:mpa-effect)

```{r mpa-effect, include=TRUE, fig.width=width_2col, fig.asp=asp_2col, fig.cap = "Results of difference-in-difference regression estimating the regional effect of the northern Channel Island MPAs on mean total biomass densities of targeted species (difference in mean total biomass density of targeted species over time relative to expected levels using non-targeted species as a control). Grey distributions show posterior probability distribution of estimated MPA effect; red point is median estimated effect, thicker red section 50% credible interval, thinner red line 90% credible interval). Blue distributions in background show range of MPA effects produced by simulation model tuned to reflect the dynamics of the Northern Channel Island MPAs (black dashed line is median simulated value). Results are estimated in blocks of three years, including years greater than or equal to left-hand value and less than right-hand value."}

@@

The network of MPAs implemented in the Northern Channel Islands in 2003 provides an ideal opportunity to examine the regional effects of protected area networks. This analysis is to our knowledge the first empirical estimate of the regional effects of a large and iconic MPA network on a broad assemblage of targeted fin-fish. Despite the well enforced nature of these MPAs [XX is there any citation for this XX], and the presence of a rigorous scientific monitoring program, we are unable to estimate a clear effect of protection. While response ratios of targeted species have increased over time, providing evidence of effective *within MPA* protection, these response ratios are on their own an unreliable indicator of the regional conservation effects of the MPAs. Our difference-in-difference regression estimates a high probability of increasingly positive MPA effects over the first nine years of protection, only to see these estimated gains reversed from 2012 to 2017. For the most recent time period, we can detect no clear positive or negative regional effect of the Channel Island MPAs on biomass densities of targeted fin-fish (Fig.\@ref(fig:mpa-effect)).

How can we explain these results? One explanation may lie in fleet dynamics. Much of the theoretical literature on MPAs assumes that all else being equal bigger reserves produce bigger conservation gains [@white2011 and references therein]. However, these models generally simulate fleet dynamics through fishing mortality rates (e.g. concentration of fishing mortality @halpern2004). The assumption of these models is that fishers determine an amount off effort to exert, and distribute that effort outside the MPA in response to some function.

@@

While we do not have access to fine scale fishing data from the Northern Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule out a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Biogeographic differences in the distributions of targeted and non-targeted species may confound the observed effects of MPAs. Many of the non-targeted species in the Channel Islands have warm thermal affinities and have increased in numbers since the heatwave [@freedman2019]. The targeted group is made up mostly of fishes with cold-water affinities, such as members of the genus *Sebastes*. As such, we hypothesize that the recent evidence for a decline in densities of targeted species is due to environmental conditions that disproportionately affect the targeted group (and not for example due to concentrated fishing pressure outside the reserves). This hypothesis is supported by the declining trend in mean biomass densities of targeted species seen inside the MPAs themselves. If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas (Fig.\@ref(fig:raw-density)).

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, The Channel Islands would seem at face value to be an ideal location to study the regional conservation effects of protected areas. The persistently high response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs, the MPAs may still be providing protection within their borders. But, as we have shown here these response ratios are not necessarily an indicator of regional effects. The difference-in-difference strategy utilized here presents an alternative identification strategy, that while not without its own strict caveats presents some potential improvements over response-ratios as a means of estimating regional conservation effects. While we estimate a highly uncertain but overall positive effect at first, we are unable to detect a robust signal from 2012-2017. We believe that the dissapearance of increasing probabilities of positive MPA effects estimated by our model are likely driven by the marine heatwave experienced by the region beginning in 2014. After 14 years of MPA protection and following a large environmental perturbation we are left without a clear picture of the effect of the Channel Island MPA network on biomass densities of targeted fin-fish species.

>>>>

---
title: Regional Effects of Marine Protected Area Networks Are Unclear
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity called for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature is now calling for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they may benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of numerical or biomass densities inside their borders, relative to these metrics in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of the importance and challenges of estimating regional MPA effects. A network of MPAs covering 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. We are unable to find a clear effect of these same MPAs on biomass densities throughout the Channel Islands region. We use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

 year_bins <-
  seq(2003, max(pisco_data$year) + 1, by = 3)

 sim_mpa_effects <- outcomes %>% 
  filter(set == "ci",
         final_depletion > 0,
         f_v_m > .5) %>% 
  mutate(year = years_protected + 2002) %>% 
  filter(year <= max(pisco_data$year)) %>% 
  mutate(binned_year = cut(year, year_bins, include.lowest = FALSE))  %>% 
  filter(year >= 2003,
         !is.na(binned_year)) %>% 
  mutate(mpa_effect = pmin(mpa_effect,3))
```

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas (MPAs), parts of the oceans protected from human activities such as fishing, are increasingly being used to conserve and manage these ecosystems. Our study pairs theory with empirical methods to show that the regional conservation effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a large and well protected network of MPAs in the Channel Island, California, after 14 years of protection, which our simulation analysis suggests is actually to be expected. MPA practitioners must clearly communicate expectations from MPA networks, and plan monitoring and adaptations strategies that directly address the challenges of estimating regional conservation effects of protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years, with special thanks to to Kathyrn Davis Koehn and Avrey Parsons-Field. DO thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments and technical support

@@

Recent agreements to expand MPAs (most proximately, the Convention on Biological Diversity’s Strategic Plan for Biodiversity, which calls for 10% of coastal waters to be protected inside MPAs by 2020 and the International Union for Conservation of Nature call for 30% by 2030) are based on the expectation that well-designed MPAs will achieve benefits both within, and outside, their borders. Despite these assumptions, our collective scientific understanding of the regional-scale conservation and fishery impacts of current and future MPAs is surprisingly limited.

@@

As stakeholders around the world seek to use MPAs in marine resource management portfolios, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. This is particularly true in the context of a changing climate, in which "set it and forget it" management strategies are increasingly untenable; we need to be able to monitor the performance of MPAs in order to adapt them as dictated by a shifting environment. The MPAs within the Channel Islands National Marine Sanctuary, California, USA (which we will refer to as the Channel Islands from now on) provide an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 [see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs]. This MPA network has subsequently been used as a model case study in protected area design around the world [@botsford2014 and references therein].

@@

We define the regional conservation effects of MPAs as the change in mean total biomass densities of targeted finfish both inside and outside of MPAs, relative to the mean total biomass densities of targeted finfish inside and outside of MPAs that would have occurred without the MPAs (acknowledging that other outcomes such as increased biodiversity or resiliency are also important to conservation but are beyond the scope of this analysis). Consider a evenly distributed population that has 50% of its range protected by an MPA. Suppose that the MPA increase biomass densities inside the reserves by 20%, and by 0% outside the reserve. By our metric the regional conservation effect of the MPA would be 10%.

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), the age and degree of enforcement [@edgar2014],  how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through network effects [@gaines2010; @costello2010a].

@@

The MPAs in this study cover 21% of the the surface waters of the Northern Channel Islands, a region spanning roughly 90km east to west. Detailed dispersal studies are not available for all of the species covered by our study, but what information we have suggest that while adults of some of these species are likely to exhibit site fidelity [@love2011], larvae are likely distributed beyond the Channel Islands [@rassweiler2012]. Therefore, if we assume that the population of the targeted species is at least equal to the extent of the Channel Islands, we can assume that at most the MPAs cover roughly 20% of the targeted populations' ranges.

@@

```{r}
median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected> 13])
```

We integrated these network design features, life history traits, and exploitation histories into a spatially explicit bio-economic simulation model, and used this model to generate expected outcomes for the Channel Islands MPAs. Our results suggest that while a wide range of outcomes are plausible, from 0% to upwards of 200%, the a median simulated effect size for a stylized version of the Channel Islands after over ten years of protection was roughly `r percent(median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected > 10]))`; a potentially small value from the perspective of empirical detectability given the natural variability and sampling challenges of marine environments.

@@

In the case of the Channel Islands MPAs, control sites are often located within a few kilometers of an MPA, suggesting that they are susceptible to biological spillover and concentration of fishing effort resulting from the MPAs. Given these complications, we can interpret the response ratios reported in @caselle2015, and updated here, as evidence that the MPAs are providing effective protection of targeted fish biomass inside their borders, particularly in light of the general increase in response ratios over time. However, it is unclear how well these response ratio results serve as in indicator of overall regional effects of the MPAs.

To illustrate this problem, we simulated response ratio trajectories for species and MPA coverage representative of the Channel Islands (see methods and SI for details of model structure). These simulations cover the life-histories and MPA sizes seen in the Channel Islands, but vary in key unknowns such as the actual degree of fishing pressure, the timing of density dependence, and the fishing fleets response to the MPAs. For each year of protection, we paired our simulated response ratios to the estimated posterior probability distributions of the Channel Island response ratios. For example, if the mean estimated response ratio in the year 2006 is one, we found all simulations with three years of protection that had simulated response ratios near one, and then pulled out the "true" simulated regional MPA effect from each of those simulations. This provides us with a distribution of simulated regional MPA effects that could plausibly generate the types of response ratios actually observed in the Channel Islands.

The response ratio trends we observe in the data could plausibly be produced by a wide range of regional MPA effects (Fig.\@ref(fig:rr-plot)). Response ratios well over one were associated with regional MPA effects generally less than 25%, and many simulations produced large response ratios but regional MPA effects close to 0% This can occur if for example fishing pressure is only moderate, adult movement is low, larval dispersal is high, and displaced fishing effort concentrates around the border of the MPAs. The response ratio results previously published for this region do not serve as sufficient evidence for the regional conservation effects of the Channel Islands MPA network.

@@

Building off of the concepts explored in @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group. Targeted species in the Channel Islands, include commercially important fin-fish such as California sheephead (*Semicossyphus pulcher*), and copper (*Sebastes caurinus*) and blue (*Sebastes mystinus*) rockfish. Each of these targeted species was the subject of prior bio-economic modeling related to the effects of MPAs in southern California [@costello2010a;@rassweiler2014]. It should be noted though that our analysis omits species from important invertebrate fisheries including red urchin (*Mesocentrotus franciscanus*) and spiny lobster (*Panulirus interruptus*). Included non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*) (See Table.S2 for a complete list of species). We used a Bayesian difference-in-difference regression to estimate any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation [see @larsen2019 for an explanation of difference-in-difference estimators in ecological settings].  The result of this regression is, conditional on the assumptions of the model, an estimate of the effect of the MPAs on the mean total biomass densities of targeted species throughout the Channel Islands.

@@

Consistent with MPA theory, over the first three years of implementation (2003-2006), the effects of the MPAs are unclear, with support for a small negative effect to a substantially positive effect, with much higher probability of a small positive effect (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2003,2006]"]) - 1, accuracy = 1)`). Over the next six years the model estimates greater probabilities of an increasingly positive MPA effect, peaking in 2009-2011 with a median estimate of MPA effect of a `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)` increase in mean total biomass density of targeted species (90% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2009,2012]"]) - 1, accuracy = 1)`. These empirical estimates are in line with (though in the upper range of) the outcomes that our simulation model suggests are plausible given the rough characteristics of the northern Channel Island MPAs. While this concordance between theory and empirics is by no means proof of the robustness of our results, it is an important line of evidence. However, in the subsequent years the trend reverses itself, and for the years 2015-2017 we once again see no clear effect of the MPAs (median estimated effect `r percent(exp(did_coefs$estimate[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`,
90\% credible interval `r percent(exp(did_coefs$lower[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`
- `r percent(exp(did_coefs$upper[did_coefs$term == "targeted:year_bins(2015,2018]"]) - 1, accuracy = 1)`.) Fig.\@ref(fig:mpa-effect)

```{r mpa-effect, include=TRUE, fig.width=width_2col, fig.asp=asp_2col, fig.cap = "Results of difference-in-difference regression estimating the regional effect of the Channel Island MPAs on mean total biomass densities of targeted species (difference in mean total biomass density of targeted species over time relative to expected levels using non-targeted species as a control). Grey distributions show posterior probability distribution of estimated MPA effect; red point is median estimated effect, thicker red section 50% credible interval, thinner red line 90% credible interval). Blue distributions in background show range of MPA effects produced by simulation model tuned to reflect the dynamics of the Channel Island MPAs (black dashed line is median simulated value). Results are estimated in blocks of three years, including years greater than or equal to left-hand value and less than right-hand value."}

@@

<!-- The network of MPAs implemented in the Northern Channel Islands in 2003 provides an ideal opportunity to examine the regional effects of protected area networks. This analysis is to our knowledge the first empirical estimate of the regional effects of a large and iconic MPA network on a broad assemblage of targeted fin-fish. Despite the well enforced nature of these MPAs, and the presence of a rigorous scientific monitoring program, we are unable to estimate a clear effect of protection. While response ratios of targeted species have increased over time, providing evidence of effective *within MPA* protection, these response ratios are on their own an unreliable indicator of the regional conservation effects of the MPAs. Our difference-in-difference regression estimates a high probability of increasingly positive MPA effects over the first nine years of protection, only to see these estimated gains reversed from 2012 to 2017. For the most recent time period, we can detect no clear positive or negative regional effect of the Channel Island MPAs on biomass densities of targeted fin-fish (Fig.\@ref(fig:mpa-effect)).  -->

How can we explain the lack of a clear regional MPA effect after over ten years of protection? One explanation may lie in fleet dynamics. Much of the theoretical literature on MPAs assumes that all else being equal bigger reserves produce bigger conservation gains [@white2011 and references therein]. However, these models generally simulate fleet dynamics through fishing mortality rates (e.g. concentration of fishing mortality @halpern2004). The assumption of these models is that fishers determine an amount off effort to exert, and distribute that effort outside the MPA in response to some function.

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule out a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Biogeographic differences in the distributions of targeted and non-targeted species may confound the observed effects of MPAs. Many of the non-targeted species in the Channel Islands have warm thermal affinities and have increased in numbers since the heatwave [@freedman2019]. The targeted group is made up mostly of fishes with cold-water affinities, such as members of the genus *Sebastes*. We hypothesize that the recent evidence for a decline in densities of targeted species is due to environmental conditions that disproportionately affect the targeted group (and not for example due to concentrated fishing pressure outside the reserves). This hypothesis is supported by the declining trend in mean biomass densities of targeted species seen inside the MPAs themselves. If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas (Fig.\@ref(fig:raw-density)).

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, The Channel Islands would seem at face value to be an ideal location to study the regional conservation effects of protected areas. The persistently high response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs, the MPAs may still be providing protection within their borders. But, as we have shown here these response ratios are not necessarily an indicator of regional effects. The difference-in-difference strategy utilized here presents an alternative identification strategy, that while not without its own strict caveats presents some potential improvements over response-ratios as a means of estimating regional conservation effects. While we estimate a highly uncertain but overall positive effect at first, we are unable to detect a robust signal from 2012-2017. We believe that the dissapearance of increasing probabilities of positive MPA effects estimated by our model are likely driven by the marine heatwave experienced by the region beginning in 2014. After 14 years of MPA protection and following a large environmental perturbation we are left without a clear picture of the effect of the Channel Island MPA network on biomass densities of targeted fin-fish species from either response ratios or the difference-in-difference model.
